\vspace{-3em} \textbf{ Mapping stellar surfaces\\ II: An interpretable Gaussian process model for light curves }

### 2021-01-29 paper2/ms.tex
finished paper2 intro

Over the past two decades, Gaussian processes (GPs) have gained traction as
a leading tool for modeling correlated signals in astronomical datasets
\citep[][]{RasmussenWilliams2005}.
In particular, GPs are commonly used to model
stellar variability in photometric time series
\citep[e.g.,][]{BrewerStello2009,Aigrain2016,Luger2016,ForemanMackey2017,Angus2018}
and radial velocity measurements \citep[e.g.,][]{Rajpaul2015,Jones2017,Perger2020}.
GPs are popular models for these applications because they allow marginalization
over a stochastic noise process specified only by a kernel describing its autocorrelation structure.
%
There are several popular open source implementations that allow
efficient evaluation of GPs, and these have been widely demonstrated
to be useful effective models for the time series when the stochastic
variability due to the star is primarily a nuisance
\citep[e.g.,][]{Ambikasaran2015,ForemanMackey2017,Gilbertson2020}.

@@

A major source of stellar variability in both light curves and radial velocity
datasets is the modulation induced by magnetically-driven surface features like starspots
rotating in and out of view. While GPs excel at marginalizing over stellar rotational
variability, they have been less useful when the goal is to make
inferences about the actual source of this variability, such as the
properties of starspots and the magnetic processes that generate them.
\textbf{While it is straightforward to derive
    posterior constraints on the hyperparameters of an effective GP model for
    observations of a star, it is not clear what those constraints actually
    tell us about the stellar surface.}
Specifically, in all but a few restricted cases, there is no \emph{first principles}
relationship between the descriptive parameters of a typical GP model
(see \S\ref{sec:gp-intro})
and the physical properties of the stellar surface that is being observed.
%
For instance, it may be tempting to interpret the GP amplitude hyperparameter
as some measure of the spot contrast or the total number of spots,
or the GP timescale hyperparameter as the
spot lifetime, but there are no guarantees these interpretations will
hold in general. After all, the choice of kernel is quite often \emph{ad hoc},
providing an \emph{effective}---as opposed to \emph{interpretable}---description
of the physics.
%
There are two important exceptions to this: asteroseismic studies, in which the
the GP hyperparameters can offer direct insight into
the behavior of complex pulsation modes and thus physical properties of the
stellar interior \citep[e.g.,][]{BrewerStello2009,ForemanMackey2017}; and
stellar rotation period studies, in which the period hyperparameter $P$ can usually
be associated with the rotation period of the star \citep[e.g.,][]{Angus2018}.%
\footnote{An exception to this is in the presence of strong differential
    rotation, in which case many periods may be present in the data, or
    when spots evolve coherently, which can also introduce weak periodicities
    in the light curve.}
%
For spot-induced variability, on the other hand,
GPs are usually used
when the variability itself is a nuisance parameter. For example, if the
goal is to constrain the properties of a transiting exoplanet or to
search for a planetary signal in a radial velocity dataset, a GP might be
used to remove (or, better yet, to marginalize over) the stellar variability
\citep[e.g.,][]{Haywood2014,Rajpaul2015,Luger2017b}.
In this case, the physics behind the variability is irrelevant, so an
effective model of this sort is generally sufficient.

\xxx{WORK IN PROGRESS:}

However, understanding the properties of stellar surfaces and starspots
in particular is a crucial step
toward understanding stellar magnetism, which plays a fundamental part
in stellar interior structure and evolution. Stellar magnetic fields
control the spin-down of stars over time, on which the field of
gyrochronology is founded \citep[][]{Barnes2001,Angus2019}. They affect
wave propagation in stellar interiors and must be properly
understood to interpret asteroseismic measurements \citep[e.g.,][]{Fuller2015}.
Strong magnetic fields are also likely the driving force behind chemical
peculiarity in Ap/Bp stars~\citep{Turcotte2003,Sikora2018}, as well as
radius inflation in M dwarfs \citep{Gough1966,Ireland2018}.
Stellar magnetohydrodynamics (MHD) is therefore an active area of
research \citep[e.g.,][]{Miesch2009,XXX}.
%
But because of the nonlinearity of the MHD equations and the vast range
of scales on which magnetic processes operate, theoretical progress
is difficult; there is still significant theoretical uncertainty over how dynamos
manifest in stars of different masses, how magnetism interacts with rotation,
and how star spots form \citep{Yadav2015,Weber2016}. Because of the
significant impacts of magnetism on every branch of stellar astrophysics,
these difficulties make new constraints from observations especially valuable.

\xxx{MITIGATING EXOPLANET SYSTEMATICS} \citep{Rackham2018}.

@@

When the goal is to learn about the stellar surface, the common approach
in the literature
has been to explicitly \emph{forward model} the stellar surface and its effect on the light curve, then
seek to constrain the surface properties by solving the inverse problem.
We discussed this approach in \citet{PaperI} \citepalias[hereafter][]{PaperI},
where we argued a unique solution to the surface map of
the star from a rotational light curve is not possible without the use of
aggressive (and often \emph{ad hoc}) priors. The degeneracies at play
make it effectively impossible for one to know the exact configuration
of starspots and other features on the surface of a star from its rotational
light curve alone.

@@

Thus, if we were able to derive robust posterior constraints
on the properties of each of the spots on a star, we could then
\emph{marginalize} (integrate) over them to infer the properties
describing the distribution of all the spots as a whole.
%In principle,
%this would work even if our posterior distributions contain
%strong degeneracies: as we will see, the statistical properties of
%starspots have a much smaller null space than the properties of
%individual spots.
We could do this in the usual way described above, by
modeling the properties of each of the spots in a posterior sampling
scheme, but also including a few
\emph{hyperparameters} controlling the distribution of those properties
across all spots (acting, therefore, as a prior term): i.e., a one-level
hierarchical model. The marginal posteriors for the hyperparameters, then,
would encode what we actually wish to know.
%
In practice, however, the degeneracies and often extreme multi-modality
of the distributions of individual spot properties would make this
quite hard (and expensive) to perform.
%
If only we could use the elegant machinery of Gaussian processes to
perform this marginalization for us!

\textbf{In this paper, we derive an exact, closed-form expression for
    the Gaussian approximation to the marginal likelihood of a light curve
    conditioned on the statistical properties of starspots.} Our Gaussian
process analytically marginalizes over the degenerate and often unknowable
distributions of properties of individual starspots, revealing the
constraints imposed on the bulk spot properties without the need to
explicitly model or sample over properties of individual spots. It inherits
the speed, ease-of-use, and all other properties of traditionally-used
GPs, with the added benefit of direct physical interpretability of its
hyperparameters.
%
In particular, \textbf{our GP is aimed at ensemble analyses of many light curves.}
As we showed in \citetalias{PaperI}, the joint information content of
the light curves of many stars can break many of the degeneracies in the
mapping problem. We encourage readers to read \citetalias{PaperI} to
better understand this and other points regarding the information theory
behind stellar rotational light curves.

@@

We would like to thank David W. Hogg, Michael Gully-Santiago, Adam Jermyn, and the
Astronomical Data Group at the Center for Computational Astrophysics for
many thought-provoking discussions that helped shape this paper.

>>>>

Over the past two decades, Gaussian processes \citep[GPs;][]{RasmussenWilliams2005} have gained traction as
a leading tool for modeling correlated signals in astronomical datasets.
In particular, GPs are commonly used to model
stellar variability in photometric time series
\citep[e.g.,][]{BrewerStello2009,Aigrain2016,Luger2016,ForemanMackey2017,Angus2018}
and radial velocity measurements \citep[e.g.,][]{Rajpaul2015,Jones2017,Perger2020}.
GPs are popular models for these applications because they allow marginalization
over a stochastic noise process specified only by a kernel describing its autocorrelation structure.
%
There are several popular open source implementations that allow
efficient evaluation of GPs, and these have been widely demonstrated
to be useful effective models for the time series when the stochastic
variability due to the star is primarily a nuisance
\citep[e.g.,][]{Ambikasaran2015,ForemanMackey2017,Gilbertson2020}.

@@

A major source of stellar variability in both light curves and radial velocity
datasets is the modulation induced by magnetically-driven surface features like starspots
rotating in and out of view. While GPs excel at marginalizing over stellar rotational
variability, they have been less useful when the goal is to make
inferences about the actual source of this variability, such as the
properties of starspots and the magnetic processes that generate them.
\textbf{While it is straightforward to derive
    posterior constraints on the hyperparameters of an effective GP model for
    observations of a star, it is not clear what those constraints actually
    tell us about the stellar surface.}
Specifically, in all but a few restricted cases, there is no \emph{first principles}
relationship between the descriptive parameters of a typical GP model
(see \S\ref{sec:gp-intro})
and the physical properties of the stellar surface that is being observed.
%
For instance, it may be tempting to interpret the GP amplitude hyperparameter
as some measure of the spot contrast or the total number of spots,
or the GP timescale hyperparameter as the
spot lifetime, but there are no guarantees these interpretations will
hold in general. After all, the choice of kernel is quite often \emph{ad hoc},
providing an \emph{effective}---as opposed to \emph{interpretable}---description
of the physics.
%
There are two important exceptions to this: asteroseismic studies, in which the
the GP hyperparameters can offer direct insight into
the behavior of complex pulsation modes and thus physical properties of the
stellar interior \citep[e.g.,][]{BrewerStello2009,ForemanMackey2017}; and
stellar rotation period studies, in which the period hyperparameter $P$ can usually
be associated with the rotation period of the star \citep[e.g.,][]{Angus2018}.%
\footnote{An exception to this is in the presence of strong differential
    rotation, in which case many periods may be present in the data, or
    when spots evolve coherently, which can also introduce weak periodicities
    in the light curve.}
%
For spot-induced variability, on the other hand,
GPs are usually used
when the variability itself is a nuisance parameter. For example, if the
goal is to constrain the properties of a transiting exoplanet or to
search for a planetary signal in a radial velocity dataset, a GP might be
used to remove (or, better yet, to marginalize over) the stellar variability
\citep[e.g.,][]{Haywood2014,Rajpaul2015,Luger2017b}.
In this case, the physics behind the variability is irrelevant, so an
effective model of this sort may be sufficient.

However, understanding the properties of stellar surfaces and starspots
in particular is a crucial step
toward understanding stellar magnetism, which plays a fundamental part
in stellar interior structure and evolution. Stellar magnetic fields
control the spin-down of stars over time, on which the field of
gyrochronology is founded \citep[][]{Barnes2001,Angus2019}. They affect
wave propagation in stellar interiors and must be properly
understood to interpret asteroseismic measurements \citep[e.g.,][]{Fuller2015}.
Strong magnetic fields are also likely the driving force behind chemical
peculiarity in Ap/Bp stars~\citep{Turcotte2003,Sikora2018}, as well as
radius inflation in M dwarfs \citep{Gough1966,Ireland2018}.
Stellar magnetohydrodynamics (MHD) is therefore an active area of
research, with many open questions \citep[e.g.,][]{Miesch2009}.
%
Because of the nonlinearity of the MHD equations and the vast range
of scales on which magnetic processes operate,
there is still significant theoretical uncertainty concerning how dynamos
operate in stars of different masses, how magnetic fields affect stellar rotation,
and how star spots form \citep{Yadav2015,Weber2016}.
Observational constraints on starspots and other magnetically-controlled
surface features are therefore extremely valuable to understanding various problems in
stellar astrophysics.

Moreover, even when the stellar signal is considered a nuisance, a
physically-driven variability model may be a better choice than an
effective model in some cases, particularly when the signal of interest is small
compared to the systematics. A specific example of this is in transmission
spectroscopy of transiting exoplanets, where the contribution from
unocculted spots and faculae to the spectrum
can be an order of magnitude larger than that of the planet atmosphere
\citep{Rackham2018}. In this case, failure to explicitly model the effect of starspots
can lead to spurious features in the planet spectrum.
%
A similar situation arises in extreme precision radial velocity (EPRV)
searches for planets, where the stellar signal can be orders of magnitude
larger than the planetary signal. While effective models of
variability have often been successful at disentangling the planetary
and stellar contributions \xxx{\citep{XXX}}, these models can struggle
when the (a priori unknown) orbital period of the planet
is close to an alias of the rotational period of the star \xxx{\citep{XXX}}.
In this case, a physically-driven model of variability would likely
perform better.

@@

When the goal is to learn about the stellar surface, the common approach
in the literature has not been to use GPs, but to explicitly
\emph{forward model} the surface. Such a model allows one to
compute a stellar light curve or spectral timeseries conditioned
on certain surface properties, a procedure that must then be inverted
in order to constrain the surface given a dataset.
We discussed this approach for rotational light curves of stars
in \citet{PaperI} \citepalias[hereafter][]{PaperI},
where we argued a unique solution to the surface map of
the star is not possible without the use of
aggressive (and often \emph{ad hoc}) priors. \textbf{The degeneracies at play
    make it effectively impossible for one to know the exact configuration
    of starspots and other features on the surface of a star from its rotational
    light curve alone.}

@@

Thus, if it were possible to derive robust posterior constraints
on the properties of each of the spots on a star, we could then
\emph{marginalize} (integrate) over them to infer the properties
describing the distribution of all the spots as a whole.
%In principle,
%this would work even if our posterior distributions contain
%strong degeneracies: as we will see, the statistical properties of
%starspots have a much smaller null space than the properties of
%individual spots.
We could do this using the forward model approach described above, by
modeling the properties of each of the spots and computing the
corresponding light curves. Then, we could solve the ``inverse'' problem
via a posterior sampling scheme, such as Markov Chain Monte Carlo (MCMC),
while including a few
\emph{hyperparameters} controlling the distribution of those properties
across all spots: i.e., a one-level hierarchical model.
The marginal posteriors for the hyperparameters, then,
would encode what we actually wish to know.
%
In practice, however, the degeneracies and often extreme multi-modality
of the distributions of individual spot properties would make this
quite hard (and expensive) to perform.
%
If only we could use the elegant machinery of Gaussian processes to
perform this marginalization for us!

\textbf{In this paper, we derive an exact, closed-form expression for
    the Gaussian approximation to the marginal likelihood of a light curve
    conditioned on the statistical properties of starspots,} which allows us
define an interpretable Gaussian process for stellar light curves.
%
Our GP analytically marginalizes over the degenerate and often unknowable
distributions of properties of individual starspots, revealing the
constraints imposed on the bulk spot properties without the need to
explicitly model or sample over properties of individual spots. It inherits
the speed, ease-of-use, and all other properties of traditionally-used
GPs, with the added benefit of direct physical interpretability of its
hyperparameters.
%
In particular, \textbf{our GP is aimed at ensemble analyses of many light curves.}
As we showed in \citetalias{PaperI}, the joint information content of
the light curves of many stars can be harnessed to constrain
statistical properties of the surfaces of those stars, even in the
presence of degeneracies that preclude knowledge about the surfaces
of individual stars.
We encourage readers to read \citetalias{PaperI} to
better understand this and other points regarding the information theory
behind stellar rotational light curves.

@@

We would like to thank David W. Hogg, Michael Gully-Santiago, Adam Jermyn,
Megan Bedell, and the
Astronomical Data Group at the Center for Computational Astrophysics for
many thought-provoking discussions that helped shape this paper.
